---
title: "SCCAN tutorial and tests"
bibliography: REFERENCES.bib
output: pdf_document
---
```{r global options, include=FALSE,echo=TRUE, message=FALSE, warning=FALSE}
library( knitr )
library( randomForest )
library( visreg )
library(pheatmap)
library(png)
invisible(suppressMessages(library(ANTsR)))
rootdir<-"./"
if ( ! file.exists(rootdir) ) {
  dirs<-list.dirs(path = path.expand("~"), full.names = TRUE, recursive = TRUE)
  rootdir<-dirs[ grep('sccanTutorial',dirs)[1] ]
  rootdir<-path.expand(rootdir)
}
opts_knit$set(root.dir=rootdir)
its=10 # could go up to 20 or 40
verb=0 # verbosity
copt=0 # 0 for most applications, 1 for priors
smth=0 # set to 0.5 to help avoid overfitting
```

# Introduction

## Brief summary
Sparse canonical correlation analysis for neuroimaging (SCCAN) is a general purpose
tool for "two-sided" multiple regression.  This allows one to symmetrically
compare one matrix of data to another and find linear relationships between
them in a low-dimensional space.  SCCAN derives from classic canonical correlation
analysis and also relates to singular value decomposition.  To handle data with $p>>n$,
SCCAN uses high-dimensional regularization methods common in $\ell_1$ regression
and spatial regularization to help ensure the biological plausibility of statistical
maps in medical imaging.  This problem is a difficult optimization ($np$-hard)
and, to improve solution interpetability and stability, SCCAN allows one to
to use prior knowledge to constrain the solution space.  This tutorial is based
on dimensionality reduction ideas outlined in the
Eigenanatomy (@Dhillon2014) and SCCAN (@Avants2014) papers.

# Examples

Perhaps the best way to understand how to use SCCAN is by running example data.
The example data below consists of measurements from cortical gray matter and
measurements from a diverse cognitive battery, the Philadelphia Brief Assessment
of Cognition.  The code and data below are at
[https://github.com/stnava/sccanTutorial](https://github.com/stnava/sccanTutorial)
and depend on [ANTsR commit](https://github.com/stnava/ANTsR/commit/2dc5060de3146401f261918172636be95584d2f6)
or more recent.

## Read example data
We read in some neuroimaging and cognitive data below.
```{r readata,fig.width=8, fig.height=4, echo=TRUE, message=FALSE, warning=FALSE }
data(aal,package='ANTsR')
gfnl<-list.files(path=rootdir, pattern = glob2rx("pbac*mha"),
  full.names = T,recursive = T)
ptrainimg<-as.matrix(antsImageRead(gfnl[2],2))
ptestimg<-as.matrix(antsImageRead(gfnl[1],2))
gfnl<-list.files(path=rootdir, pattern = "gmask.nii.gz",
  full.names = T,recursive = T)
mask<-antsImageRead( gfnl[1], 3 )
afnl<-list.files(path=rootdir, pattern = "aal.nii.gz",
  full.names = T,recursive = T)
aalimg<-antsImageRead( afnl[1], 3 )
f1<-list.files(path =rootdir, pattern = "pbac_train_cog.csv",
  recursive=TRUE, full.names = TRUE, include.dirs=TRUE )
f2<-list.files(path = rootdir, pattern = "pbac_test_cog.csv",
  recursive=TRUE, full.names = TRUE )
ptraincog<-read.csv(f1)
ptestcog<-read.csv(f2)
```
We already divided the dataset into two different groups - one for testing and
one for training.  

## Sparse regression

Use SCCAN to find brain regions relating to age.  In this case, sparse CCA
acts like a sparse regression.  We impose a "cluster
threshold" regularization to prevent isolated voxels from appearing in
the solution.  We will also compare the results in training with that
in testing as a function of spareseness.  This type of approach can
be useful in parameter selection i.e. in choosing the optimization
criterion based on training data.
```{r sparreg,fig.width=8, fig.height=4, echo=TRUE, message=FALSE, warning=FALSE}
agemat<-matrix( ptraincog$age, ncol=1)
paramsearch<-c(1:10)/(100.0)
paramsearchcorrs<-rep(0,length(paramsearch))
paramsearchpreds<-rep(0,length(paramsearch))
ct<-1
nvox = sum( mask >= 0.5 )
imat = abs( t(replicate(2, rnorm(ncol(ptrainimg)))) )
for ( sp in paramsearch ) {
  imats=imat
  imats[1, order(imat[1,])[1:round(nvox*(1-abs(sp)))] ]=0
  imats[2, order(imat[2,])[1:round(nvox*(1-abs(sp)))] ]=0
  initdf<-initializeEigenanatomy( imats, mask=mask )
  ageresult<-sparseDecom2( inmatrix=list(ptrainimg,agemat), its=its,
    sparseness=c(sp,0.9), inmask=c(mask,NA),nvecs=2, mycoption=copt,
    cthresh=c(1000,0), ell1=0.1,
    smooth=smth, verbose=verb ) # initializationList = initdf$initlist,
  sccanimg1=matrixToImages( t(ageresult$eig1), mask )[[1]]
  # validate in test data
  agepred<-ptrainimg %*% (ageresult$eig1)
  paramsearchcorrs[ct]<-abs( cor( agepred[,1],  ptraincog$age ) )
  agepred<-ptestimg %*% ageresult$eig1
  paramsearchpreds[ct]<-abs( cor( agepred[,1],  ptestcog$age ) )
  print(paste(sp,paramsearchcorrs[ct],paramsearchpreds[ct]))
  ct<-ct+1
  }
mydf<-data.frame( sparseness=paramsearch, trainCorrs=paramsearchcorrs,
  testCorrs=paramsearchpreds )
mdl1<-lm( trainCorrs ~ sparseness+I(sparseness^2), data=mydf )
mdl2<-lm( testCorrs ~ sparseness+I(sparseness^2), data=mydf )
visreg(mdl1)
visreg(mdl2)
```

## SCCAN with prior initialization

Use SCCAN to find brain regions relating to a battery of tests
that measure language-related cognitive function.
We initialize SCCAN with left hemisphere regions.  In this case,
the initialization controls the sparseness parameters for each
eigenvector.  The parameter *sp* will be overridden by the priors,
thereby enabling a "per eigenvector" sparseness value.
```{r sparreginit,fig.width=8, fig.height=4, echo=TRUE, message=FALSE, warning=FALSE}
langmat<-cbind(  ptraincog$speech_adj, ptraincog$writing_adj,
                 ptraincog$semantic_adj, ptraincog$reading_adj,
                 ptraincog$naming_adj )
colnames(langmat)<-c("speech","writing","semantic","reading","naming")
langmat2<-cbind( ptestcog$speech_adj, ptestcog$writing_adj,
                 ptestcog$semantic_adj, ptestcog$reading_adj,
                 ptestcog$naming_adj )
colnames(langmat2)<-colnames(langmat)
labels<-c(13,81,39,79)
print(aal$label_name[labels])
initmat<-matrix( rep(0,sum(mask==1)*length(labels)), nrow=length(labels) )
# fill the matrix with the aal region locations
for ( i in 1:length(labels) ) {
  vec<-( aalimg[ mask == 1 ] == labels[i] )
  initmat[i,]<-as.numeric( vec )
}
ccainit<-initializeEigenanatomy( initmat, mask )
pwsearch<-c(50,25,10)
langfn<-rep("",length(pwsearch))
langfn2<-rep("",length(pwsearch))
ct<-1
for ( pw in pwsearch ) {
langresult<-sparseDecom2( inmatrix=list(ptrainimg,langmat), its=its,
  mycoption=copt,
  sparseness=c(sp,-0.5), inmask=c(mask,NA),nvecs=length(labels), cthresh=c(1000,0), verbose=1,
  initializationList=ccainit$initlist, priorWeight=pw/100, smooth=smth,
  perms=0 )
ccamat = langresult$eig1
langpred<-ptrainimg %*% (ccamat)
colnames(langpred)<-paste("GM",c(1:ncol(langpred)),sep='')
cogpred<-langmat %*% ( langresult$eig2  )
bestpred<-which.max(abs(diag(cor(langpred,cogpred))))
mydf<-data.frame( cogpred, langpred )
myform<-as.formula( paste("Variate00",bestpred-1,"~GM1+GM2+GM3+GM4",sep='') )
mdltrain<-lm( myform, data=mydf )
langpred<-ptestimg %*% (ccamat)
colnames(langpred)<-paste("GM",c(1:ncol(langpred)),sep='')
cogpred<-langmat2 %*% data.matrix( langresult$eig2 )
mydf<-data.frame( cogpred, langpred )
print(cor.test( mydf[,bestpred] ,predict(mdltrain,newdata=mydf)))
for ( i in 1:length(labels) )
  print( paste( "Dice: ",aal$label_name[labels[i]],
         sum( abs(t(ccamat)[i,]) > 0 & initmat[i,] > 0 ) /
         sum( abs(t(ccamat)[i,]) > 0 | initmat[i,] > 0 ) ) )
langimgs = matrixToImages( t(langresult$eig1) , mask )
for ( x in langimgs ) {
  x[ mask == 1 ]<-abs( x[ mask == 1 ] )
  x[ mask == 1 ]<-x[ mask == 1 ]/max( x[ mask == 1 ] )
}
mycolors<-c("red","green","blue","yellow")
langfn[ct]<-paste(rootdir,'/figures/langSCCANRegression',pw,'.jpg',sep='')
langfn2[ct]<-paste(rootdir,'/figures/langSCCANRegression',pw,'.png',sep='')
plot( mask, langimgs, color.overlay=mycolors,
      outname=langfn[ct], axis=1 )
# cnt<-getCentroids( ntwkimage, clustparam = 100 )
brain<-renderSurfaceFunction( surfimg =list( aalimg ) , alphasurf=0.1 ,
  funcimg=langimgs, smoothsval=1.5, smoothfval=0, mycol=mycolors )
id<-par3d("userMatrix")
rid<-rotate3d( id , -pi/2, 1, 0, 0 )
rid2<-rotate3d( id , pi/2, 0, 0, 1 )
rid3<-rotate3d( id , -pi/2, 0, 0, 1 )
par3d(userMatrix = id )
dd<-make3ViewPNG(  rid, id, rid2, paste(rootdir,'/figures/langSCCANRegression',pw,sep='') )
par3d(userMatrix = id )
ct<-ct+1
}
```
Strong use of prior

![Strong prior](`r langfn[1]`)

![Strong prior 3D](`r langfn2[1]`)

Medium use of prior

![Medium prior](`r langfn[2]`)

![Medium prior 3D](`r langfn2[2]`)

Weak use of prior

![Weak prior](`r langfn[3]`)

![Weak prior 3D](`r langfn2[3]`)

## Identifying the anatomical network

The best results are initialized by the prior but, in the end,
drift away from that initialization.  Where in the brain do the solution
vectors end up?  We write a quick function to answer this question.


```{r aallabel, message=FALSE, warning=FALSE}
reportAnatomy<-function( eigIn, maskIn, wt=0.3 )
  {
  data('aal',package='ANTsR')
  ccaanat<-list()
  for ( img in eigIn ) {
    nzind<-abs(img[ maskIn == 1 ]) > 0
    aalvals<-aalimg[ maskIn == 1 ][ nzind ]
    ccaanat<-lappend( ccaanat, aalvals )
  }
  ccaanat<-unlist( ccaanat )
  anatcount<-hist(ccaanat,breaks=0:100, plot = F)$count
  anatcount[ anatcount < wt*max(anatcount) ]<-0
  anatcount<-which( anatcount > 0 )
  return( toString(aal$label_name[anatcount] ) )
  }
ccaaal<-reportAnatomy( matrixToImages( t(langresult$eig1),mask) , mask )
```
The SCCAN predictors include: `r ccaaal`.

How good were our original hypothetical regions as predictors?


## Associating classes to SCCAN predictors

Recalling: CCA maximizes $PearsonCorrelation( XW^T, ZY^T )$,
where $X$ and $Z$ are data matrices, we can study matrix $Y$ (or $W$)
which contrasts or combines columns of the associated data matrix.
In this example, $Y$ operates on the cognition/design matrix.

```{r sccanpredictorclass, message=FALSE, warning=FALSE}
rownames(langresult$eig2)<-colnames(langmat)
temp<-(langresult$eig2)
temp[ abs(langresult$eig2) < 0.0001 ]<-0
```

```{r sccanpredictorclass2, message=FALSE, warning=FALSE}
pheatmap(temp)
```


## Sparse regression with nuisance variables

Often, we want to control for the presence of nuisance variables.
As usual, there are several options:  (1) control after you do
dimensionality reduction;  (2) orthogonalize the predictors.
(3) Use alternative SCCAN formulations (e.g. set `mycoption` to 0 or 2).
Let's try the first 2 choices as they are more traditional.

```{r covar1, fig.width=8, fig.height=4, echo=TRUE, message=FALSE, warning=FALSE}
# 1. control for age and mmse after the dimensionality reduction
langresult<-sparseDecom2( inmatrix=list(ptrainimg,langmat), its=its,
  mycoption=copt,
  sparseness=c(sp,-0.9), inmask=c(mask,NA),nvecs=length(labels), cthresh=c(1000,0),
  initializationList=ccainit$initlist, priorWeight=pw/100, smooth=smth )
ccamat = langresult$eig1
langpred<-ptrainimg %*% (ccamat)
colnames(langpred)<-paste("GM",c(1:ncol(langpred)),sep='')
cogpred<-langmat %*% data.matrix( langresult$eig2 )
bestpred<-which.max(abs(diag(cor(langpred,cogpred))))
mydf<-data.frame( cogpred, langpred, mmse=ptraincog$mmse,age=ptraincog$age)
myform<-as.formula( paste("Variate00",bestpred-1,
  "~GM1+GM2+GM3+GM4+mmse+age",sep='') )
mdltrain<-lm( myform, data=mydf )
print(summary(mdltrain))
langpred2<-ptestimg %*% (ccamat)
colnames(langpred2)<-paste("GM",c(1:ncol(langpred)),sep='')
cogpred2<-langmat2 %*% data.matrix( langresult$eig2 )
mydf<-data.frame( cogpred2, langpred2,mmse=ptestcog$mmse,age=ptestcog$age )
print(cor.test( mydf[,bestpred] ,predict(mdltrain,newdata=mydf)))
```

Now, the second option.
```{r covar2, fig.width=8, fig.height=4, echo=TRUE, message=FALSE, warning=FALSE}
# 2. orthogonalize the matrices against mmse and education
rlangmat<-residuals(lm(langmat~ptraincog$mmse+ptraincog$age))
rptrainimg<-residuals(lm(ptrainimg~ptraincog$mmse+ptraincog$age))
langresult<-sparseDecom2( inmatrix=list(rptrainimg,rlangmat), its=its,
  mycoption=copt,
  sparseness=c(sp,-0.9), inmask=c(mask,NA),nvecs=length(labels), cthresh=c(1000,0),
  initializationList=ccainit$initlist, priorWeight=pw/100, smooth=smth )
ccamat = langresult$eig1
langpred<-ptrainimg %*% (ccamat)
colnames(langpred)<-paste("GM",c(1:ncol(langpred)),sep='')
cogpred<-langmat %*% data.matrix( langresult$eig2 )
bestpred<-which.max(abs(diag(cor(langpred,cogpred))))
mydf<-data.frame( cogpred, langpred, mmse=ptraincog$mmse,age=ptraincog$age)
myform<-as.formula( paste("Variate00",bestpred-1,
  "~GM1+GM2+GM3+GM4+mmse+age",sep='') )
mdltrain<-lm( myform, data=mydf )
print(summary(mdltrain))
langpred2<-ptestimg %*% (ccamat)
colnames(langpred2)<-paste("GM",c(1:ncol(langpred)),sep='')
cogpred2<-langmat2 %*% data.matrix( langresult$eig2 )
mydf<-data.frame( cogpred2, langpred2,mmse=ptestcog$mmse,age=ptestcog$age )
print(cor.test( mydf[,bestpred] ,predict(mdltrain,newdata=mydf)))
```


## Predicting the full cognitive battery from the neuroimaging data

Try to predict all the demographic variability from the imaging data.
We use `mycoption 0` to try to reduce correlation in low-dimensional
space.  This enforces a new SCCAN constraint (not previously reported).
```{r fullpred, fig.width=8, fig.height=4, echo=TRUE, message=FALSE, warning=FALSE}
nv<-11
nfn<-rep("",nv)
cognames<-rep("",nv)
cogmat<-data.matrix(ptraincog)
rcogmat<-residuals( lm( data.matrix(ptraincog) ~ ptraincog$mmse + ptraincog$age ) )
rptrainimg<-residuals( lm( ptrainimg ~ ptraincog$mmse ) )
batt<-sparseDecom2( inmatrix=list(rptrainimg,rcogmat), its=its,
  sparseness=c(0.015, -0.05), inmask=c(mask,NA), nvecs=nv,
  cthresh=c(1000,0), smooth=smth, mycoption=copt )
ccamat = batt$eig1
gvars<-paste("GM",c(1:ncol(ccamat)),sep='',collapse='+')
```

Now let's use our previously developed reporting capabilities.
```{r fullreport, fig.width=8, fig.height=4, echo=TRUE, message=FALSE, warning=FALSE}
render<-TRUE
batteig = matrixToImages( t(batt$eig1), mask )
for ( bestpred in 1:ncol(ccamat)) {
  battpred<-ptrainimg %*% (ccamat)
  colnames(battpred)<-paste("GM",c(1:ncol(battpred)),sep='')
  cogpred<-( rcogmat %*% data.matrix( batt$eig2 ) )[,bestpred]
  mydf<-data.frame( cogpred, battpred, age=ptraincog$age, mmse=ptraincog$mmse )
  myform<-as.formula( paste("cogpred~",gvars,'+age+mmse',sep='') )
  mdltrain<-lm( myform, data=mydf )
  mdlinterp<-bigLMStats( mdltrain )
  battpred<-ptestimg %*% (ccamat)
  colnames(battpred)<-paste("GM",c(1:ncol(battpred)),sep='')
  cogpred<-(data.matrix(ptestcog) %*% data.matrix( batt$eig2 ))[,bestpred]
  mydf<-data.frame( cogpred, battpred, age=ptestcog$age, mmse=ptestcog$mmse  )
  cat(paste("Eig",bestpred,"is related to:\n"))
  mycog<-colnames(ptraincog)[ abs(batt$eig2[,bestpred]) > 0 ]
  cat( mycog )
  cat("\nwith weights\n")
  cat( abs(batt$eig2[,bestpred])[ abs(batt$eig2[,bestpred]) > 0 ])
  cat(paste("\nwith predictive correlation:",
    cor( cogpred,predict(mdltrain,newdata=mydf))))
  cat("\nAnatomy:")
  for ( x in which.min(p.adjust(mdlinterp$beta.pval)) )  {
    myanat<-reportAnatomy( list( batteig[[x]]) , mask , 0.5 )
    cat(myanat)
    if ( render ) {
    vizimg<-abs( antsImageClone( batteig[[x]] ) )
    brain<-renderSurfaceFunction( surfimg =list( aalimg ) , alphasurf=0.1 ,
      funcimg=list(vizimg), smoothsval = 1.5 )
    id<-par3d("userMatrix")
    rid<-rotate3d(  id , -pi/2, 1, 0, 0 )
    rid2<-rotate3d( id ,  pi/2, 0, 0, 1 )
    rid3<-rotate3d( id , -pi/2, 0, 0, 1 )
    par3d(userMatrix = id )
    ofn<-paste(rootdir,'/figures/battery',bestpred,sep='')
    nfn[ bestpred ]<-paste(ofn,'.png',sep='')
    cognames[ bestpred ]<-paste(mycog,collapse='+')
    dd<-make3ViewPNG(  rid, id, rid2, ofn )
    par3d(userMatrix = id )
    }
    cat("\n")
  }
  cat("\n")
}
```

Anatomy related with `r cognames[1]`

![Select results](`r nfn[1]`)

Anatomy related with `r cognames[2]`

![Select results](`r nfn[2]`)

Anatomy related with `r cognames[3]`

![Select results](`r nfn[3]`)

Anatomy related with `r cognames[4]`

![Select results](`r nfn[4]`)

Anatomy related with `r cognames[5]`

![Select results](`r nfn[5]`)

Anatomy related with `r cognames[6]`

![Select results](`r nfn[6]`)

Anatomy related with `r cognames[7]`

![Select results](`r nfn[7]`)

Anatomy related with `r cognames[8]`

![Select results](`r nfn[8]`)

Anatomy related with `r cognames[9]`

![Select results](`r nfn[9]`)

Anatomy related with `r cognames[10]`

![Select results](`r nfn[10]`)

Anatomy related with `r cognames[11]`

![Select results](`r nfn[11]`)


Can the neuroimaging data predict the full cognitive battery?

```{r mvarpred,fig.width=8, fig.height=4, echo=TRUE, message=FALSE, warning=FALSE}
# use cca to transform cortical signal to the cognitive battery
batt2<-sparseDecom2( inmatrix=list(rptrainimg,rcogmat), its=its,
  sparseness=c(0.02, -0.9), inmask=c(mask,NA), nvecs=nv,
  cthresh=c(1000,0), smooth=smth, mycoption=copt )
ccamat = batt2$eig1
predictedBattery<-data.frame( vox=ptrainimg %*% (ccamat) %*% t(batt2$eig2) )
print(diag(cor(cogmat,predictedBattery)))
predictedBattery<-data.frame( vox=ptestimg %*% (ccamat) %*% t(batt2$eig2) )
print(diag(cor(data.matrix(ptestcog),predictedBattery)))
qv<-rep(NA,ncol(ptestcog) )
for ( i in 1:ncol(ptestcog) ) {
 qv[i]<-cor.test(data.matrix(ptestcog)[,i],predictedBattery[,i])$p.value
 ttl<-paste(  colnames(ptestcog)[i],
      cor(data.matrix(ptestcog)[,i],predictedBattery[,i]) )
 mdl<-data.frame( realCog=data.matrix(ptestcog)[,i],
                  predCog=predictedBattery[,i] )
 mylm<-lm( predCog ~ realCog , data=mdl )
 visreg( mylm , main=ttl)
 Sys.sleep(1)
}
qv[ is.na(qv) ]<-1
qv<-p.adjust(qv,method='BH')
sum(qv<0.05)/length(qv)
```
The following univariate columns may be predicted
using SCCAN multivariate mapping: `r colnames(ptestcog)[ qv <= 0.05 ] `.
